---
title: "Intro and regression refresher"
subtitle: "EDUC 643: General Linear Model I"
author: "TBD"
#date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  xaringan::moon_reader:
    css: ['default','new.css']
    # self_contained: true
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false

---

```{R, setup, include = F}
library(pacman)
p_load(here, tidyverse, ggplot2, xaringan, knitr, kableExtra, foreign)

i_am("slides/EDUC643_1_intro.rmd")

# Define color
red_pink <- "#e64173"
turquoise <- "#20B2AA"
orange <- "#FFA500"
red <- "#fb6107"
blue <- "#3b3b9a"
green <- "#8bb174"
grey_light <- "grey70"
grey_mid <- "grey50"
grey_dark <- "grey20"
purple <- "#6A5ACD"
slate <- "#314f4f"
# Dark slate grey: #314f4f

# Knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  fig.height = 6.75,
  fig.width = 10.5,
  warning = F,
  message = F
)
opts_chunk$set(dev = "svg")
options(device = function(file, width, height) {
  svg(tempfile(), width = width, height = height)
})
options(knitr.table.format = "html")

```
# Roadmap
<font size = "3">
.pull-left[
1. A refresher on foundations of the General Linear Model
 - Review of bivariate regression
 - Coefficient and model-level inference
 - Correlation and causality
 
2. Assumptions and diagnostics
 - Residuals, studentized, standardized
 - Normality of residuals, homoscedasticity, linearity and independence
 - Diagnostics

3. Multiple regression
 - Concept of statistical adjustment (control)
 - Variance decomposition in MR
 - Measuring effect size in MR
 - Multi-collinearity
]

.pull-right[
4. Categorical predictors
 - Coding dummy variables
 - ANOVA as a special regression model with one categorical predictor
 - Adding a continuous predictor, ANCOVA as a special regression model with a categorical predictor and one or many continuous predictors
 - Variance decomposition when categorical predictors are in the model

5. Interactions and non-linearity
 - Intro to the concept of interaction in MR models
 - Interpreting the interaction between a categorical predictor and continuous predictor
 - Interpreting the interaction between two continuous predictors

6. Model building
]
</font>

---
class: middle, inverse

# Bivariate regression

---
<<<<<<< HEAD
## A motivating question

```{r, echo=T}
do <- read.spss(here::here("data/male_do_eating.sav"), to.data.frame=T)
summary(do$EDEQ_restraint)
summary(do$BMI)
```

---
## Bivariate relationships
```{r, echo=F}
ggplot(do, aes(x=BMI, y=EDEQ_restraint)) + geom_point()
```
---
=======
>>>>>>> 476a8018f846dc4c6b72c607adb0cf80932ae1f0
class: middle, inverse

# Regression inference

---
class: middle, inverse

# Correlation and causality

