---
title: "Describing Variables and Relationships"
author: enter_your_name
date: "2024-01-08"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
#change eval = TRUE and keep everything else the same
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      messages = FALSE,
                      eval = FALSE)
```

**GOALS:**

1. Introduce .Rmd file
2. Revise code to import and format dataset 
3. Revise code to describe variables and relationships

```{r load libraries}
library(tidyverse)
library(modelsummary)
```

### Reading and inspecting dataset 

```{r read dataset}
# we will use the dataset for the final project for EDUC 641 - ah01.csv
# make a project and recommended folder for this class - save this .Rmd and the dataset in the folder, then read the dataset

# replace the dots with code
health <- read.csv("...")

# check the number of rows and columns and the current type of column

#replace the dots with code
str(...)
```

*Q1. How many rows and columns are there? What is the current type of the following variables: id, mentor, education, gpa, gpa_3? What is the correct scale of measurement of each variable (nominal, ordinal, interval, or ratio)? Which are continuous and categorical variables?*

**Answer here**

### Changing variable types

```{r convert to factor}
# change id, mentor, and gpa_3 to factor variables
# one is fone for you

health$id <- factor(health$id) # we don't specify levels and labels here because they are too many. R will use the exisiting names as default levels

# replace the dots with code
health$mentor <- factor(health$mentor,
                     levels = c(0,1),
                     labels = c("...", "..."))

health$gpa_3 <- ...
```

### Describing variables

```{r descriving categorical variables}
# report the number and proportion of students that had a mentor and those having a score of 3 or more on their GPA.

table(health$mentor)
prop.table(table(health$mentor))

# replace the dots with code
...(health$gpa_3)
...(...(health$...))
```

*Q2. How many students had a mentor? What proportion of students had a GPA score of 3 or more?*

**Asnwer here**

```{r descriving continuous variables}
# report the mean, median and standard deviation of the education variable.

mean(health$education, na.rm = TRUE)
median(...$..., na.rm = TRUE)
sd(...$..., na.rm = TRUE)

# do the same for the gpa variable

mean(health$gpa, na.rm = TRUE)
median(...$..., na.rm = TRUE)
sd(...$..., na.rm = TRUE)
```

*Q3. Report the summary statistics for the education variable, i.e., the total number of years a student recieved formal education?*

**Asnwer here**

### Plotting variables and relationships 

```{r bar plot for categorical variable and relationships}
# make a bar plot to visualize the number of students who had a gpa score of 3 or more

# replace the dots with code
ggplot(data = ..., aes(x = ...))+
  geom_bar(fill = "...") +
  labs(x = "...",
       y = "...")+
  theme_minimal()

# visually inspect the relationship between gpa_3 and mentor

ggplot(data = ..., aes(x = mentor, fill = gpa_3)) +
  geom_bar(position = "dodge") +
  labs(x = "...",
       y = "...",
       fill = "...")+
  theme_minimal()
```

*Q4. Comment about the visual relationship between mentor and gpa_3.*

**Answer here**

```{r distribution of continuous variable and relationships}
# make a historgram to visualize the distribution of the education variable

# replace the dots with code
ggplot(data = ..., aes(x = education))+
  geom_histogram(fill = "...", bins = 7) +
  labs(x = "...",
       y = "...")+
  theme_minimal()

# visually inspect the relationship between education and mentor

ggplot(data = ..., aes(y = education, fill = mentor)) +
  geom_boxplot(show.legend = F)+
  labs(x = "...",
       y = "...",
       fill = "...")+
  theme_minimal()
```

*Q5. Comment about the visual relationship between mentor and education*

**Answer here**

```{r bivariate relationships}
# make a scatter plot to visualize the relationship between education and gpa

# replace the dots with code
ggplot(data = ..., aes(x = ..., y = ...))+
  geom_point()+
  labs(x = "...",
       y = "...")+
  theme_minimal()
```

*Q6. Comment about the visual relationship between education and gpa.*

**Answer here**

